---
title: "ICU Mortality Prediction — Machine Learning Pipeline"
format: html
---

# ICU Mortality Prediction — Machine Learning Pipeline

This **self-contained** R report builds and evaluates predictive models for
in-hospital mortality using OMOP CDM data queried via `sql_query()`.

Steps:

1. Cohort extraction (SQL)
2. Feature engineering (SQL + R)
3. Data preparation & cleaning
4. Train/test split
5. Logistic regression (baseline)
6. Decision tree (rpart)
7. Model comparison & evaluation
8. Calibration analysis
9. Feature importance
10. Clinical interpretation & summary

**Requirements:** An active database connection with OMOP CDM tables
(person, visit_occurrence, death, measurement, concept).

> `sql_query(sql)` is automatically available in Linkr R notebooks.
> It queries the active DuckDB connection and returns a data.frame.

```{r setup}
set.seed(42)
```

## 1. Cohort Extraction

We create the cohort entirely via SQL — eligible visits (≥ 24h stay),
mortality status, and demographics.

```{r cohort-views}
# Eligible visits: stays >= 24 hours
sql_query("
    CREATE OR REPLACE VIEW eligible_visits AS
    SELECT
        v.visit_occurrence_id, v.person_id,
        v.visit_start_date,
        v.visit_start_datetime::TIMESTAMP AS visit_start_datetime,
        v.visit_end_date,
        v.visit_end_datetime::TIMESTAMP AS visit_end_datetime,
        EXTRACT(EPOCH FROM (v.visit_end_datetime::TIMESTAMP
            - v.visit_start_datetime::TIMESTAMP)) / 3600 AS los_hours
    FROM visit_occurrence v
    WHERE EXTRACT(EPOCH FROM (v.visit_end_datetime::TIMESTAMP
        - v.visit_start_datetime::TIMESTAMP)) / 3600 >= 24
")

# In-hospital mortality
sql_query("
    CREATE OR REPLACE VIEW visit_mortality AS
    SELECT ev.*,
        CASE WHEN d.death_date IS NOT NULL
             AND d.death_date BETWEEN ev.visit_start_date
                                   AND ev.visit_end_date + INTERVAL '1 day'
             THEN 1 ELSE 0 END AS in_hospital_death
    FROM eligible_visits ev
    LEFT JOIN death d ON ev.person_id = d.person_id
")

# Final cohort with demographics
sql_query("
    CREATE OR REPLACE VIEW cohort AS
    SELECT vm.visit_occurrence_id, vm.person_id, vm.visit_start_datetime,
        EXTRACT(YEAR FROM vm.visit_start_date) - p.year_of_birth AS age,
        p.gender_source_value AS sex, vm.los_hours, vm.in_hospital_death
    FROM visit_mortality vm
    JOIN person p ON vm.person_id = p.person_id
    WHERE EXISTS (
        SELECT 1 FROM measurement m
        WHERE m.visit_occurrence_id = vm.visit_occurrence_id
          AND m.value_as_number IS NOT NULL
          AND m.measurement_datetime::TIMESTAMP >= vm.visit_start_datetime
          AND m.measurement_datetime::TIMESTAMP
              <= vm.visit_start_datetime + INTERVAL '24 hours'
    )
")
```

```{r cohort-summary}
summary_df <- sql_query("
    SELECT COUNT(*) AS n_visits, COUNT(DISTINCT person_id) AS n_patients,
        SUM(in_hospital_death) AS n_deaths,
        ROUND(100.0 * SUM(in_hospital_death) / COUNT(*), 1) AS mortality_pct,
        ROUND(AVG(age), 1) AS mean_age,
        ROUND(AVG(los_hours), 1) AS mean_los_hours
    FROM cohort
")
cat("Cohort:\n")
print(summary_df)
```

## 2. Feature Engineering

Extract vital signs, laboratory values, and GCS from the first 24 hours,
then aggregate into a wide-format feature matrix.

```{r measurement-config}
# Concept IDs → column prefixes
vitals <- c(
    "3027018" = "hr",        "3004249" = "sbp",       "3012888" = "dbp",
    "3027598" = "mbp",       "3024171" = "resp_rate",  "40762499" = "spo2",
    "3020891" = "temp"
)
labs <- c(
    "3000963" = "hemoglobin", "3023314" = "hematocrit", "3024929" = "platelets",
    "3003282" = "wbc",        "3019550" = "sodium",     "3023103" = "potassium",
    "3014576" = "chloride",   "3016293" = "bicarbonate","3016723" = "creatinine",
    "3013682" = "bun",        "3004501" = "glucose",    "3037278" = "anion_gap",
    "3015377" = "calcium",    "3012095" = "magnesium",  "3011904" = "phosphate"
)
neuro <- c(
    "3016335" = "gcs_eye", "3009094" = "gcs_verbal", "3008223" = "gcs_motor"
)
all_measurements <- c(vitals, labs, neuro)
```

```{r extract-measurements}
concept_ids <- paste(names(all_measurements), collapse = ", ")

measurements_h24 <- sql_query(paste0("
    SELECT m.visit_occurrence_id, m.measurement_concept_id,
           m.value_as_number,
           m.measurement_datetime::TIMESTAMP AS measurement_datetime,
           c.visit_start_datetime
    FROM measurement m
    JOIN cohort c ON m.visit_occurrence_id = c.visit_occurrence_id
    WHERE m.measurement_concept_id IN (", concept_ids, ")
      AND m.value_as_number IS NOT NULL
      AND m.measurement_datetime::TIMESTAMP >= c.visit_start_datetime
      AND m.measurement_datetime::TIMESTAMP
          <= c.visit_start_datetime + INTERVAL '24 hours'
"))

cat(sprintf("Measurements in H0-H24: %d rows\n", nrow(measurements_h24)))
```

```{r aggregate-features}
# Map concept IDs to feature names
measurements_h24$feature <- all_measurements[as.character(measurements_h24$measurement_concept_id)]

vitals_ids  <- as.integer(names(vitals))
labs_ids    <- as.integer(names(labs))
neuro_ids   <- as.integer(names(neuro))

# Aggregate per visit x feature
agg_list <- list()
for (vid in unique(measurements_h24$visit_occurrence_id)) {
    sub <- measurements_h24[measurements_h24$visit_occurrence_id == vid, ]
    for (feat in unique(sub$feature)) {
        rows <- sub[sub$feature == feat, ]
        cid <- rows$measurement_concept_id[1]
        vals <- rows$value_as_number[order(rows$measurement_datetime)]

        if (cid %in% vitals_ids) {
            agg_list[[length(agg_list) + 1]] <- data.frame(
                visit_occurrence_id = vid,
                col = paste0(feat, c("_mean", "_min", "_max")),
                val = c(mean(vals), min(vals), max(vals)),
                stringsAsFactors = FALSE
            )
        } else if (cid %in% neuro_ids) {
            agg_list[[length(agg_list) + 1]] <- data.frame(
                visit_occurrence_id = vid,
                col = paste0(feat, "_min"), val = min(vals),
                stringsAsFactors = FALSE
            )
        } else if (cid %in% labs_ids) {
            agg_list[[length(agg_list) + 1]] <- data.frame(
                visit_occurrence_id = vid,
                col = paste0(feat, "_first"), val = vals[1],
                stringsAsFactors = FALSE
            )
        }
    }
}
agg_df <- do.call(rbind, agg_list)
cat(sprintf("Aggregated: %d rows, %d features\n",
    nrow(agg_df), length(unique(agg_df$col))))
```

```{r pivot-and-merge}
# Pivot to wide format
wide <- reshape(agg_df, idvar = "visit_occurrence_id", timevar = "col",
                direction = "wide", v.names = "val")
names(wide) <- sub("^val\\.", "", names(wide))

# Merge with cohort demographics
cohort_df <- sql_query("
    SELECT visit_occurrence_id, person_id, age, sex, los_hours, in_hospital_death
    FROM cohort
")
df <- merge(cohort_df, wide, by = "visit_occurrence_id", all.x = TRUE)

cat(sprintf("Dataset: %d rows x %d columns\n", nrow(df), ncol(df)))
cat(sprintf("Deaths: %d / %d (%.1f%%)\n",
    sum(df$in_hospital_death), nrow(df),
    100 * mean(df$in_hospital_death)))
```

## 3. Data Preparation

```{r feature-selection}
id_cols     <- c("visit_occurrence_id", "person_id")
target_col  <- "in_hospital_death"
exclude     <- c(id_cols, target_col, "sex", "los_hours")

feature_cols <- setdiff(names(df), exclude)
feature_cols <- feature_cols[sapply(df[feature_cols], is.numeric)]

# Exclude features with > 30% missing
missing_pct <- sapply(df[feature_cols], function(x) mean(is.na(x)))
usable <- names(missing_pct[missing_pct < 0.30])
excluded <- names(missing_pct[missing_pct >= 0.30])

cat(sprintf("Total features: %d\n", length(feature_cols)))
cat(sprintf("Usable (<30%% missing): %d\n", length(usable)))
if (length(excluded) > 0) {
    cat(sprintf("Excluded: %s\n", paste(excluded, collapse = ", ")))
}
```

```{r model-matrix}
# Build model matrix with sex encoding and median imputation
model_df <- df[, c(target_col, usable), drop = FALSE]
model_df$sex_male <- as.integer(df$sex == "M")

for (col in usable) {
    nas <- is.na(model_df[[col]])
    if (any(nas)) {
        model_df[[col]][nas] <- median(model_df[[col]], na.rm = TRUE)
    }
}

feature_names <- c(usable, "sex_male")
cat(sprintf("Model matrix: %d x %d\n", nrow(model_df), length(feature_names)))
```

## 4. Train/Test Split

```{r train-test-split}
# Stratified 75/25 split
pos_idx <- which(model_df[[target_col]] == 1)
neg_idx <- which(model_df[[target_col]] == 0)

train_pos <- sample(pos_idx, round(0.75 * length(pos_idx)))
train_neg <- sample(neg_idx, round(0.75 * length(neg_idx)))
train_idx <- sort(c(train_pos, train_neg))
test_idx  <- setdiff(seq_len(nrow(model_df)), train_idx)

train_df <- model_df[train_idx, ]
test_df  <- model_df[test_idx, ]

cat(sprintf("Train: %d (%.1f%% mortality)\n",
    nrow(train_df), 100 * mean(train_df[[target_col]])))
cat(sprintf("Test:  %d (%.1f%% mortality)\n",
    nrow(test_df), 100 * mean(test_df[[target_col]])))
```

```{r split-comparison}
# Compare feature distributions between train and test
cat(sprintf("\n%-25s %12s %12s %8s\n", "Feature", "Train mean", "Test mean", "Diff"))
cat(paste(rep("-", 60), collapse = ""), "\n")
for (col in head(feature_names, 15)) {
    tr <- mean(train_df[[col]], na.rm = TRUE)
    te <- mean(test_df[[col]], na.rm = TRUE)
    cat(sprintf("%-25s %12.2f %12.2f %8.2f\n", col, tr, te, abs(tr - te)))
}
```

## 5. Logistic Regression

```{r logistic-regression}
formula <- as.formula(paste(target_col, "~",
    paste(feature_names, collapse = " + ")))
lr_model <- glm(formula, data = train_df, family = binomial)

cat("Model coefficients (top 15 by |coefficient|):\n")
coefs <- coef(lr_model)[-1]  # exclude intercept
coef_df <- data.frame(
    Feature = names(coefs),
    Coefficient = unname(coefs),
    OddsRatio = exp(unname(coefs)),
    stringsAsFactors = FALSE
)
coef_df <- coef_df[order(-abs(coef_df$Coefficient)), ]
print(head(coef_df, 15), row.names = FALSE)
```

```{r lr-predictions}
lr_proba_train <- predict(lr_model, train_df, type = "response")
lr_proba_test  <- predict(lr_model, test_df, type = "response")
lr_pred_test   <- as.integer(lr_proba_test >= 0.5)
```

## 6. Decision Tree

```{r rpart-model}
library(rpart)

tree_model <- rpart(formula, data = train_df, method = "class",
    control = rpart.control(cp = 0.01, maxdepth = 5, minsplit = 20))

tree_proba_test <- predict(tree_model, test_df, type = "prob")[, "1"]
tree_pred_test  <- as.integer(tree_proba_test >= 0.5)

cat("Decision tree structure:\n")
print(tree_model)
```

## 7. Model Evaluation

```{r roc-helpers}
# Manual ROC + AUC (base R, no external packages)
compute_roc <- function(y_true, y_prob) {
    thresholds <- sort(unique(c(0, y_prob, 1)), decreasing = TRUE)
    tpr <- fpr <- numeric(length(thresholds))
    n_pos <- sum(y_true == 1)
    n_neg <- sum(y_true == 0)
    for (i in seq_along(thresholds)) {
        pred <- as.integer(y_prob >= thresholds[i])
        tpr[i] <- sum(pred == 1 & y_true == 1) / max(n_pos, 1)
        fpr[i] <- sum(pred == 1 & y_true == 0) / max(n_neg, 1)
    }
    ord <- order(fpr, tpr)
    list(fpr = fpr[ord], tpr = tpr[ord], thresholds = thresholds[ord])
}

compute_auc <- function(roc) {
    abs(sum(diff(roc$fpr) * (head(roc$tpr, -1) + tail(roc$tpr, -1)) / 2))
}

compute_metrics <- function(y_true, y_pred, y_prob) {
    cm <- table(Predicted = y_pred, Actual = y_true)
    tp <- if ("1" %in% rownames(cm) && "1" %in% colnames(cm)) cm["1", "1"] else 0
    tn <- if ("0" %in% rownames(cm) && "0" %in% colnames(cm)) cm["0", "0"] else 0
    fp <- if ("1" %in% rownames(cm) && "0" %in% colnames(cm)) cm["1", "0"] else 0
    fn <- if ("0" %in% rownames(cm) && "1" %in% colnames(cm)) cm["0", "1"] else 0
    roc <- compute_roc(y_true, y_prob)
    list(
        auc     = compute_auc(roc),
        sens    = tp / max(tp + fn, 1),
        spec    = tn / max(tn + fp, 1),
        ppv     = tp / max(tp + fp, 1),
        npv     = tn / max(tn + fn, 1),
        acc     = (tp + tn) / max(tp + tn + fp + fn, 1),
        brier   = mean((y_prob - y_true)^2),
        roc     = roc,
        cm      = cm
    )
}
```

```{r model-metrics}
y_test <- test_df[[target_col]]

lr_metrics   <- compute_metrics(y_test, lr_pred_test, lr_proba_test)
tree_metrics <- compute_metrics(y_test, tree_pred_test, tree_proba_test)

# Summary table
comparison <- data.frame(
    Model       = c("Logistic Regression", "Decision Tree"),
    AUC         = round(c(lr_metrics$auc, tree_metrics$auc), 3),
    Brier       = round(c(lr_metrics$brier, tree_metrics$brier), 3),
    Sensitivity = round(c(lr_metrics$sens, tree_metrics$sens), 3),
    Specificity = round(c(lr_metrics$spec, tree_metrics$spec), 3),
    PPV         = round(c(lr_metrics$ppv, tree_metrics$ppv), 3),
    NPV         = round(c(lr_metrics$npv, tree_metrics$npv), 3),
    Accuracy    = round(c(lr_metrics$acc, tree_metrics$acc), 3),
    stringsAsFactors = FALSE
)

cat("Model Comparison:\n")
print(comparison, row.names = FALSE)
```

```{r roc-curves}
# ROC curves
par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1))

# ROC plot
plot(lr_metrics$roc$fpr, lr_metrics$roc$tpr,
     type = "l", col = "#4C78A8", lwd = 2,
     xlab = "False Positive Rate", ylab = "True Positive Rate",
     main = "ROC Curve")
lines(tree_metrics$roc$fpr, tree_metrics$roc$tpr,
      col = "#E45756", lwd = 2)
abline(0, 1, lty = 2, col = "gray50")
legend("bottomright", cex = 0.8,
    legend = c(
        sprintf("Logistic (AUC=%.3f)", lr_metrics$auc),
        sprintf("Tree (AUC=%.3f)", tree_metrics$auc)
    ),
    col = c("#4C78A8", "#E45756"), lwd = 2)

# Confusion matrix (logistic regression)
cm <- lr_metrics$cm
image(1:2, 1:2, t(cm)[, 2:1], col = heat.colors(12),
      xlab = "Predicted", ylab = "Actual", main = "Confusion Matrix (LR)",
      axes = FALSE)
axis(1, at = 1:2, labels = c("Alive", "Dead"))
axis(2, at = 1:2, labels = c("Dead", "Alive"))
for (i in 1:2) for (j in 1:2) {
    text(i, 3 - j, cm[j, i], cex = 1.5, font = 2)
}
box()
```

## 8. Calibration

```{r calibration}
# Calibration: observed vs predicted in bins
calibration_data <- function(y_true, y_prob, n_bins = 10) {
    breaks <- seq(0, 1, length.out = n_bins + 1)
    bins <- cut(y_prob, breaks, include.lowest = TRUE)
    pred_mean <- tapply(y_prob, bins, mean)
    obs_mean  <- tapply(y_true, bins, mean)
    counts    <- tapply(y_true, bins, length)
    data.frame(
        predicted = pred_mean, observed = obs_mean, n = counts,
        stringsAsFactors = FALSE
    )
}

par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1))

for (i in seq_along(list(lr_proba_test, tree_proba_test))) {
    proba <- list(lr_proba_test, tree_proba_test)[[i]]
    name  <- c("Logistic Regression", "Decision Tree")[i]
    color <- c("#4C78A8", "#E45756")[i]

    cal <- calibration_data(y_test, proba)
    cal <- cal[!is.na(cal$predicted), ]

    plot(cal$predicted, cal$observed,
         type = "b", pch = 16, col = color, lwd = 2,
         xlim = c(0, 1), ylim = c(0, 1),
         xlab = "Predicted probability", ylab = "Observed fraction",
         main = sprintf("Calibration — %s", name))
    abline(0, 1, lty = 2, col = "gray50")
}
```

## 9. Feature Importance

```{r lr-importance}
# Logistic regression: standardized coefficients
sds <- sapply(train_df[feature_names], sd, na.rm = TRUE)
std_coefs <- coefs * sds[names(coefs)]
importance_df <- data.frame(
    Feature = names(std_coefs),
    StdCoef = unname(std_coefs),
    AbsStdCoef = abs(unname(std_coefs)),
    stringsAsFactors = FALSE
)
importance_df <- importance_df[order(-importance_df$AbsStdCoef), ]

top_n <- min(20, nrow(importance_df))
top <- importance_df[1:top_n, ]

par(mfrow = c(1, 2), mar = c(4.5, 10, 3, 1))

# Standardized coefficients
barplot(rev(top$StdCoef), names.arg = rev(top$Feature),
        horiz = TRUE, las = 1, col = ifelse(rev(top$StdCoef) > 0, "#E45756", "#4C78A8"),
        main = "LR: Standardized Coefficients",
        xlab = "Standardized coefficient", cex.names = 0.7)
abline(v = 0, lty = 2)
```

```{r tree-importance}
# Decision tree: variable importance
tree_imp <- tree_model$variable.importance
if (!is.null(tree_imp) && length(tree_imp) > 0) {
    tree_imp_df <- data.frame(
        Feature = names(tree_imp),
        Importance = unname(tree_imp),
        stringsAsFactors = FALSE
    )
    tree_imp_df <- tree_imp_df[order(-tree_imp_df$Importance), ]
    top_t <- tree_imp_df[1:min(20, nrow(tree_imp_df)), ]

    par(mar = c(4.5, 10, 3, 1))
    barplot(rev(top_t$Importance), names.arg = rev(top_t$Feature),
            horiz = TRUE, las = 1, col = "#72B7B2",
            main = "Decision Tree: Variable Importance",
            xlab = "Importance", cex.names = 0.7)
}
```

## 10. Threshold Analysis

```{r threshold-analysis}
# Evaluate metrics across thresholds
thresholds <- seq(0.05, 0.95, by = 0.05)
threshold_results <- data.frame(
    Threshold = thresholds,
    Sensitivity = numeric(length(thresholds)),
    Specificity = numeric(length(thresholds)),
    PPV = numeric(length(thresholds)),
    F1 = numeric(length(thresholds)),
    stringsAsFactors = FALSE
)

for (i in seq_along(thresholds)) {
    pred <- as.integer(lr_proba_test >= thresholds[i])
    tp <- sum(pred == 1 & y_test == 1)
    tn <- sum(pred == 0 & y_test == 0)
    fp <- sum(pred == 1 & y_test == 0)
    fn <- sum(pred == 0 & y_test == 1)
    sens <- tp / max(tp + fn, 1)
    spec <- tn / max(tn + fp, 1)
    ppv  <- tp / max(tp + fp, 1)
    f1   <- 2 * sens * ppv / max(sens + ppv, 1e-9)
    threshold_results[i, 2:5] <- c(sens, spec, ppv, f1)
}

par(mfrow = c(1, 1), mar = c(4.5, 4.5, 3, 1))
plot(thresholds, threshold_results$Sensitivity,
     type = "l", col = "#4C78A8", lwd = 2, ylim = c(0, 1),
     xlab = "Threshold", ylab = "Metric",
     main = "Threshold Analysis (Logistic Regression)")
lines(thresholds, threshold_results$Specificity, col = "#E45756", lwd = 2)
lines(thresholds, threshold_results$PPV, col = "#72B7B2", lwd = 2)
lines(thresholds, threshold_results$F1, col = "#F58518", lwd = 2, lty = 2)
legend("right", cex = 0.8,
    legend = c("Sensitivity", "Specificity", "PPV", "F1"),
    col = c("#4C78A8", "#E45756", "#72B7B2", "#F58518"),
    lwd = 2, lty = c(1, 1, 1, 2))

# Optimal threshold (max F1)
best <- which.max(threshold_results$F1)
cat(sprintf("\nOptimal threshold (max F1): %.2f\n", thresholds[best]))
cat(sprintf("  Sensitivity: %.1f%%\n", 100 * threshold_results$Sensitivity[best]))
cat(sprintf("  Specificity: %.1f%%\n", 100 * threshold_results$Specificity[best]))
cat(sprintf("  PPV:         %.1f%%\n", 100 * threshold_results$PPV[best]))
cat(sprintf("  F1:          %.3f\n", threshold_results$F1[best]))
```

## 11. Summary

```{r summary}
cat(paste(rep("=", 70), collapse = ""), "\n")
cat("STUDY SUMMARY — IN-HOSPITAL MORTALITY PREDICTION\n")
cat(paste(rep("=", 70), collapse = ""), "\n")

cat(sprintf("
COHORT
  Visits:     %d
  Patients:   %d
  Deaths:     %d (%.1f%%)
  Age:        %.0f +/- %.0f years

DATA
  Features:   %d (%d clinical + sex)
  Train/Test: %d / %d (75/25 stratified split)
  Missing:    Median imputation (features with >30%% missing excluded)

RESULTS
  Logistic Regression
    AUC:  %.3f  |  Brier: %.3f
    Sens: %.1f%%  |  Spec: %.1f%%

  Decision Tree
    AUC:  %.3f  |  Brier: %.3f
    Sens: %.1f%%  |  Spec: %.1f%%

TOP PREDICTORS (by |standardized coefficient|)
",
    nrow(df), length(unique(df$person_id)),
    sum(df$in_hospital_death), 100 * mean(df$in_hospital_death),
    mean(df$age), sd(df$age),
    length(feature_names), length(usable),
    nrow(train_df), nrow(test_df),
    lr_metrics$auc, lr_metrics$brier,
    100 * lr_metrics$sens, 100 * lr_metrics$spec,
    tree_metrics$auc, tree_metrics$brier,
    100 * tree_metrics$sens, 100 * tree_metrics$spec
))

for (i in 1:min(10, nrow(importance_df))) {
    cat(sprintf("  %-30s %+.4f\n",
        importance_df$Feature[i], importance_df$StdCoef[i]))
}

cat("
LIMITATIONS
  - Single-center data (MIMIC-IV, Beth Israel Deaconess)
  - H0-H24 measurements only (early prediction window)
  - Median imputation for missing values
  - No external validation cohort
  - No temporal validation (train on earlier, test on later)
  - Potential survivorship bias (patients who die within 24h excluded)

NEXT STEPS
  - External validation on eICU or another OMOP CDM database
  - Temporal validation split
  - Multiple imputation instead of median
  - Additional features: diagnoses, procedures, medications
  - Ensemble methods (random forest, gradient boosting)
  - Prospective evaluation
")
```
